An Experimental Evaluation of Hybrid Querying on Vectors (EA\&B)

### 2025-07-30 VLDB2026/main.tex
1

	\begin{figure*}
		\centering
		\setlength{\belowcaptionskip}{-0.cm}
		\includegraphics[width=0.92\textwidth]{figures/hybrid ANNS.pdf}
		\caption{Hybrid Query Example}

@@

		\setlength{\abovecaptionskip}{0.1cm}
		\setlength{\belowcaptionskip}{-0.2cm}		
		\caption{Graph Index and IVF Index}

@@

	\textcolor{violet}{To investigate the time and space overhead of different algorithms,} we evaluate \textcolor{orange}{4} key metrics in the single-attribute scenario: index construction time, index size, \textcolor{orange}{build peak memory and search peak memory}. Notably, PASE on GIST1M and Enron is omitted, as it supports only data with dimensionality less than 512. \textcolor{orange}{We use 64 threads to build the index for the large-scale Deep100M dataset, and 32 threads for the other six datasets. Notably, the VBASE and PASE algorithms only support single-threaded construction.}

@@

\textit{\textbf{Index construction time.}}
%\textcolor{orange}{Among the regular-scale datasets shown in the left part of Figure~\ref{fig:build_time_comparison_query1}, the methods with longest build times are mainly PASE, VBASE, and ACORN-$\gamma$. PASE and VBASE only support single-threaded builds. PASE is the slowest because its on-demand loading and caching strategies frequently trigger disk I/O and page scheduling, which seriously affects efficiency. VBASE uses full-memory builds and is relatively faster. ACORN-$\gamma$  has a high computational overhead because it needs to expand the neighbor list and evaluate a large number of candidates. Its build time on some datasets even exceeds that of single-threaded VBASE. }
\textcolor{orange}{On the regular-scale datasets (Figure~\ref{fig:build_time_comparison_query1}, left), among the algorithms limited to single-threaded construction, PASE has the longest build time. This is because its on-demand loading and caching strategies frequently trigger disk I/O and page scheduling, which seriously affects efficiency. In contrast, VBASE uses full-memory builds and is therefore comparatively faster. Among the multi-threaded algorithms, ACORN-$\gamma$ is the slowest because it needs to expand the neighbor list and evaluate a large number of candidates. Its build time on some datasets even exceeds that of the single-threaded VBASE. }

\textcolor{orange}{On the large-scale dataset Deep100M(Figure~\ref{fig:build_time_comparison_query1}, right), in addition to the three methods already mentioned, we also observe that algorithms based on Vamana (FilteredDiskANN, StitchedVamana and UNG) exhibit significantly increased construction time. This is because Vamana adopts a two-pass graph construction and optimization strategy (greedy search + robust pruning), resulting in extremely high computational overhead. In contrast, the construction time of IVF-based algorithms grows much more slowly with dataset size.This is because their shard-based design and high parallel efficiency, making them more scalable and advantageous in large-scale scenarios.}

@@

	\textit{\textbf{Attribute Distribution.}} To evaluate performance across different attribute distributions, we generate base and query attribute sets with 4 representative distributions—long-tailed, normal, power-law, and uniform—across the 6 datasets described in Section~4.1.1. We evaluate all algorithms under these distributions on each dataset. The performance differences observed for the same algorithm across different datasets are relatively minor. Thus, due to space constraints, we present only the results on SIFT1M.

	As illustrated in Figure~\ref{fig:exp_3_1}, the evaluated algorithms exhibit varying degrees of sensitivity to attribute distribution shifts. \textcolor{cyan}{The proximity of the 4 curves in the figure reflects algorithm sensitivity to distributional variations. When the curves converge closely, the algorithm demonstrates strong resilience against attribute distribution shifts.} Specifically, ACORN, Milvus, Puck, CAPS, StitchedVamana, and UNG demonstrate strong robustness to such shifts. In contrast, Faiss, FilteredVamana, NHQ, VBASE, and PASE show notable sensitivity.

@@

	\textit{\textbf{Single-Attribute Selectivity.}}
%	In hybrid queries, varying attribute selectivity (AS) can significantly impact computational cost and query efficiency, as the algorithm must efficiently identify data points matching specific attributes within large datasets. AS is defined as the proportion of data points sharing a given attribute value. For example, AS of 1\% indicates that only 1\% of the dataset meets the given attribute. 
%	
%	To investigate this effect, we evaluate 4 AS settings (1\%, 25\%, 50\%, and 75\%) on the SIFT1M dataset using single-thread execution. 
%	We compare the query performance of attribute filtering algorithms across these settings, as shown in Figure~\ref{fig:exp_5_1_1_SingleLabel_1thread}.
%	Based on the query performance trends of different algorithms under varying AS, we categorize the algorithms into 3 groups:
%	
%	Experimental results show that UNG achieves the best performance across all AS settings, especially at extremely low AS (1\%). 
%	This advantage stems from its pre-filtering strategy. Pre-filtering strategy significantly reduces the ANN search space by narrowing down candidates prior to vector computation. In contrast, VBASE performs the worst at 1\% AS due to its post-filtering strategy. Post-filtering strategy first retrieves a large number of candidates and then applies filtering, incurring substantial computational overhead.

>>>>

		\includegraphics[width=0.92\textwidth]{figures/hybrid ANNS.pdf}
		\caption{Hybrid Query Example}

@@

		\caption{Graph Index and IVF Index}

@@

	\textcolor{violet}{To investigate the time and space overhead of different algorithms,} we evaluate \textcolor{orange}{4} key metrics in the single-attribute scenario: index construction time, index size, \textcolor{orange}{build peak memory and search peak memory}. Notably, PASE on GIST1M and Enron is omitted, as it supports only data with dimensionality less than 512. \textcolor{orange}{Except for algorithms that only support single-threaded construction, all other algorithms are built using 32 threads on the six datasets shown on the left of Figure~\ref{fig:build_index_comparison}, and 64 threads on the large-scale Deep100M dataset, shown on the right.}

@@

\textit{\textbf{Index construction time.}}
\textcolor{orange}{Among the regular-scale datasets shown in the left part of Figure~\ref{fig:build_time_comparison_query1}, the methods with longest build times are mainly PASE, VBASE, and ACORN-$\gamma$. PASE and VBASE only support single-threaded builds. PASE is the slowest because its on-demand loading and caching strategies frequently trigger disk I/O and page scheduling, which seriously affects efficiency. VBASE uses full-memory builds and is relatively faster. ACORN-$\gamma$  has a high computational overhead because it needs to expand the neighbor list and evaluate a large number of candidates. Its build time on some datasets even exceeds that of single-threaded VBASE. }

\textcolor{orange}{On the large-scale dataset Deep100M, besides the three methods already mentioned, we also observe that algorithms based on Vamana (such as FilteredDiskANN and UNG) exhibit significantly increased construction time. Vamana adopts a two-pass graph construction and optimization strategy (greedy search + robust pruning), resulting in extremely high computational overhead. In contrast, the construction time of IVF-based algorithms grows much more slowly with dataset size, thanks to their shard-based design and high parallel efficiency, making them more scalable and advantageous in large-scale scenarios.}

@@

	\textit{\textbf{Attribute Distribution.}} To evaluate the impact of attribute distribution, we generate base and query attributes \textcolor{orange}{on the SIFT dataset using four representative distributions: long-tail, normal, power-law, and uniform.}
%	with four representative distributions—long-tail, normal, power-law, and uniform—across six datasets. 
%	Since performance differences across datasets are small, we report results on SIFT1M due to space limits.
	As illustrated in Figure~\ref{fig:exp_3_1}, the evaluated algorithms exhibit varying degrees of sensitivity to attribute distribution shifts. \textcolor{cyan}{The proximity of the 4 curves in the figure reflects algorithm sensitivity to distributional variations. When the curves converge closely, the algorithm demonstrates strong resilience against attribute distribution shifts.} 
%	Specifically, ACORN, Milvus, Puck, CAPS, StitchedVamana, and UNG demonstrate strong robustness to such shifts. In contrast, Faiss, FilteredVamana, NHQ, VBASE, and PASE show notable sensitivity.

@@

	\textit{\textbf{Single-Attribute Selectivity.}}
	In hybrid queries, varying attribute selectivity (AS) can significantly impact computational cost and query efficiency.
%	, as the algorithm must efficiently identify data points matching specific attributes within large datasets. 
	AS is defined as the proportion of data points sharing a given attribute value. For example, AS of 1\% indicates that only 1\% of the dataset meets the given attribute. 
	To investigate this effect, we evaluate 4 AS settings (1\%, 25\%, 50\%, and 75\%) on the SIFT1M dataset using single-thread execution. 
%	We compare the query performance of attribute filtering algorithms across these settings, as shown in Figure~\ref{fig:exp_5_1_1_SingleLabel_1thread}.
	Based on the query performance trends of different algorithms under varying AS, we categorize the algorithms into 3 groups:
%	
%	Experimental results show that UNG achieves the best performance across all AS settings, especially at extremely low AS (1\%). 
%	This advantage stems from its pre-filtering strategy. Pre-filtering strategy significantly reduces the ANN search space by narrowing down candidates prior to vector computation. In contrast, VBASE performs the worst at 1\% AS due to its post-filtering strategy. Post-filtering strategy first retrieves a large number of candidates and then applies filtering, incurring substantial computational overhead.
